Naturalness in U(1) extensions of the MSSM

### 2015-01-20 paper/e6ssm-tuning-paper.tex
More typo fixing and rephrasings for clarity.

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models and explaining why
% DH:: I strongly suspect that they are invalid - but running scan
% to check this currently
the heavier Higgs values are less well populated [can't be obtained? -
are we sure those scattered points are valid?]  than in the other
models.  Therefore the fine tuining behaviour in this model is closer
to that of the MSSM, though the LHC can still constrain the most
natural scenarions from just placing tougher limits on the $Z^\prime$
mass .

@@

Finally we want to emphasise that while in Fig.~\ref{Fig:e6ssmvsmssm}
the E$_6$SSM looks more fine tuned than the MSSM this depends on the
cutoff scale.  Indeed in \cite{Athron:2013ipa} a constrained version
of the E$_6$SSM, with the cutoff at the GUT scale, is considered and
there the cE$_6$SSM was found to be less tuned than the cMSSM. Since a
125 GeV Higgs can be achieved in the E$_6$SSM with ligher stops, then
if the cutoff is large, the larger stop masses of the MSSM can make
that model more fine tuned due to large RGE effects.

To further illustrate this point we looked at how the tuning compares
for low tuning benchmarks in the MSSM and E6SSM when the highscale
boundary condition, $M_X$ is raised. These benchmarks are defined in
table \ref{} and the results are shown in
Fig.~\ref{Fig:BMs-varyMX}. Since the behaviour is a quite complicated
we now discuss these in detail as it provides some insight into the
many differences in the tuning between the two models.

In the top left panel one can see that the MSSM tuning (green)
steadily climbs as the cutoff scale is increased, as one would expect
when the tuning originates from large soft masses entering from the
RGEs. The panel on the top right confirms this, showing that the
largest tuning contributions come from $\Delta_{A_t}$ and
$\Delta_{m_{H_u}^2}$ with the former being the larger sensitivity
until $M_X\approx 10^8$ GeV at which point $\Delta_{m_{H_u}^2}$ takes
over, leading to the small kink in overall tuning which can be seen in
the green curve in the top left panel. In thus case we have chosen a
point with large mixing which is known to reduce the MSSM tuning.  We
found this does not eliminate the tuning as there is still a strong
sensitivity to $A_t$, but we did find that increasing $A_t$ lead less
fine tuning overall for the points we looked at.

Comparing the MSSM tuning to the E$_6$SSM tunings one can see that
which point is more fine tuned depends on the scale at which the
parameters are defined.  This illustrates that any statement about which
model is more tuned depends on the cutoff scale considered.

For E$_6$SSM BM1 the fine tuning is shown by the blue curve in top
left panel of Fig.~\ref{Fig:BMs-varyMX} and the individual sensitivities
are given in the bottom left panel. The tuning actually reduces
initially as the cutoff is is increased from $20$ TeV.  This occurs
because the tuning is initially dominated by $\Delta_\lambda$ (shown in
solid blue in the bottom left panel) that contains some terms
proportional to $M_Z^{\prime \, 2}$, which currently dominate the
tuning.  However as the cutoff scale is increased contributions from
the soft masses become more important and these actually start to
cancel the large contribution to $\Delta_\lambda$ coming from
$M_{Z^\prime}$ until $\Delta_\lambda$ passes through zero.  However at
the same time these large soft masses also cause other sensitivities
to grow, leading the overall tuning to start rising with $M_X$
once. Nonetheless due to the lower stop masses and lower mixing [I'm
  guessing slightly: Dylan I need to be able to refer to numbers
  here!]  compared to both of the other points until $M_X \approx
10^{11}$ GeV.  After this the tuning from $M_3$ starts to dominate in
this point.  Although at $M_{SUSY}$ this point as the same $M_3$ value
as the MSSM point, in the E$_6$SSM $M_3$ will be larger at $M_X$ due
to the altered RGE running from exotic matter\footnote{This altered
  RGE running is a result of the exotic matter introduced to keep the
  $U(1)$ anomaly free rather than being a direct consequence of the
  extra $U(1)$. }.  At the same time larger $M_3$ at higher scales
reduces the soft squark masses at $M_X$.  Therefore the stop mass
contributions are ameliorated, compared to the MSSM, both by allowing
lighter stops at $M_{SUSY}$ and by the modified RGE running.
Nonetheless the stops still have an impact, pushing up the sensitivity
to $m_{H_u}^2$\footnote{$m_{H_u}^2(M_{SUSY})$ receives a positive
  contribution to it's mass from $m_{H_u}^2(M_X)$ and a negative
  contribution from $m_{Q_3}^2$. If $m_{H_u}^2(M_{SUSY})$ is held
  fixed so that then as the cutoff increases the values of these
  softmasses at $M_X$ will be larger and there we be a bigger
  cancellation between them, making the sensitivity of $M_Z$ to both of
  these parameters larger.}  showing the most commonly discussed
aspect of the little hierarchy problem.

@@

\begin{figure}
\begin{center}
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_mssm_cutoff_scan_3_points_tuning_plot.png}} 
\resizebox{!}{5.5cm}{%
\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_all_tuning_data.png}} 
\\
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_all_tuning_data.png}}
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_mssm_mixing_small_M3_all_tuning_data.png}}%%
\caption{{\bf Top left panel:} Scatter plot of fine tuning vs high scale $M_X$ for the three benchmark points given in (?).  {\bf Top right panel:}  Individual sensitivites of MSSM BM plotted against the high scale $M_X$ which give an overall tuning shown by the green line in top left panel.  {\bf Bottom left panel:} Individual sensitivites of E6SSM BM1 plotted against the high scale $M_X$ which give an overall tuning shown by the blue line in top left panel.  {\bf Bottom right panel:}Individual sensitivites of E6SSM BM2 plotted against the high scale $M_X$ which give an overall tuning shown by the red line in top left panel.    }
\label{Fig:BMs-varyMX}
\end{center}
\end{figure}

>>>>

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models and explaining why
% DH:: I strongly suspect that they are invalid - but running scan
% to check this currently
the heavier Higgs values are less well populated [can't be obtained? -
are we sure those scattered points are valid?]  than in the other
models.  Therefore the fine tuining behaviour in this model is closer
to that of the MSSM, though the LHC can still constrain the most
natural scenarions from just placing tougher limits on the $Z^\prime$
mass.

@@

Finally we want to emphasise that while in Fig.~\ref{Fig:e6ssmvsmssm}
the E$_6$SSM looks more fine tuned than the MSSM this depends on the
high scale boundary, $M_X$, where the parameters are assumed to be set
by some SUSY breaking mechanism.  Indeed in \cite{Athron:2013ipa} a
constrained version of the E$_6$SSM, with the highscale boundary at
the GUT scale, is considered and there the cE$_6$SSM was found to be
less tuned than the cMSSM. Since a 125 GeV Higgs can be achieved in
the E$_6$SSM with ligher stops, then if the cutoff is large, the
larger stop masses of the MSSM can make that model more fine tuned due
to large RGE effects.

To further illustrate this point we looked at how the tuning varies
with $M_X$ for low tuning benchmarks in the MSSM and E6SSM. These
benchmarks are defined in table \ref{} and the results are shown in
Fig.~\ref{Fig:BMs-varyMX}. Since the behaviour is quite complicated we
now discuss these in detail as it provides some insight into the many
differences in the tuning between the two models.

\begin{figure}
\begin{center}
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_mssm_cutoff_scan_3_points_tuning_plot.png}} 
\resizebox{!}{5.5cm}{%
\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_all_tuning_data.png}} 
\\
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_all_tuning_data.png}}
\resizebox{!}{5.5cm}{%
\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_mssm_mixing_small_M3_all_tuning_data.png}}%%
\caption{{\it Top left panel:} Scatter plot of fine tuning vs high scale $M_X$ for the three benchmark points given in (?).  {\it Top right panel:}  Individual sensitivites of the MSSM benchmark plotted against the high scale $M_X$ which give an overall tuning shown by the green line in top left panel.  {\it Bottom left panel:} Individual sensitivites of E6SSM BM1 plotted against the high scale $M_X$ which give an overall tuning shown by the blue line in top left panel.  {\it Bottom right panel:} Individual sensitivites of E6SSM BM2 plotted against the high scale $M_X$ which give an overall tuning shown by the red line in top left panel.    }
\label{Fig:BMs-varyMX}
\end{center}
\end{figure}

In the top left panel one can see that the MSSM tuning (green)
steadily climbs as the cutoff scale is increased, as one would expect
when the tuning originates from large soft masses entering from the
RGEs. The panel on the top right confirms this, showing that the
largest tuning contributions come from $\Delta_{A_t}$ and
$\Delta_{m_{H_u}^2}$ with the former being the larger sensitivity
until $M_X\approx 10^8$ GeV at which point $\Delta_{m_{H_u}^2}$ takes
over, leading to the small kink in overall tuning which can be seen in
the green curve in the top left panel. In this case we have chosen a
point with large mixing, which is known to reduce the MSSM tuning.  We
found this does not eliminate the tuning as there is still a strong
sensitivity to $A_t$, but we did find that increasing $A_t$ lead less
fine tuning overall for the points we looked at.

Comparing the MSSM tuning to the E$_6$SSM tunings one can see that
which point is more fine tuned depends on the scale at which the
parameters are defined.  This illustrates that any statement about
which model is more tuned depends on the high scale boundary, $M_X$.

For E$_6$SSM BM1 the fine tuning is shown by the blue curve in top
left panel of Fig.~\ref{Fig:BMs-varyMX} and the individual
sensitivities are given in the bottom left panel. The tuning actually
reduces initially as the cutoff is is increased from $20$ TeV.  This
occurs because the largest sensitivity is initially $\Delta_\lambda$
(shown in solid blue in the bottom left panel).  This contains some
terms proportional to $M_Z^{\prime \, 2}$, which provide the dominant
contribution to this sensitivity at very low $M_X$.  However as $M_X$
is increased contributions from the soft masses become more important
and these actually start to cancel the large contribution to
$\Delta_\lambda$ coming from $M_{Z^\prime}$ until $\Delta_\lambda$
passes through zero.  At the same time though these large soft masses
also cause other sensitivities to grow, in particular $\Delta_{M_3}$.
The fine tuning rises with $M_X$ once $M_X \gtrsim 10^5-10^6$ GeV,
but remains lower than that of the other points, until $M_X \approx
10^{11}$ GeV.  Eventuially the $\Delta_{M_3}$ sensitivity leads to
this point being the most fine tuned of the three shown in
Fig.~\ref{Fig:BMs-varyMX}.

 Although the gluino mass and $M_3(M_{SUSY})$ have similar values to
 those in the MSSM point, in the E$_6$SSM $M_3(M_X)$ is larger
 due to the altered RGE running from exotic matter\footnote{This
   altered RGE running is a result of the exotic matter introduced to
   keep the extra $U(1)$ anomaly free. }. This is why the E$_6$SSM
 point has the much larger tuning, coming from $\Delta_{M_3}$.

Interstingly other sensitivities are suppressed by this effect since
at the same time larger $M_3$ at higher scales reduces the soft squark
masses at $M_X$.  Therefore the stop mass contributions are
ameliorated, compared to the MSSM, both by allowing lighter stops at
$M_{SUSY}$ and by the modified RGE running.  Nonetheless the stops
still do lead to $\Delta_{m_{H_u}^2}$ increasing with the cutoff
through the usual mecahnism \footnote{Wherein $m_{H_u}^2(M_{SUSY})$
  receives a positive contribution to it's mass from $m_{H_u}^2(M_X)$
  and a negative contribution from $m_{Q_3}^2(M_X)$ and
  $m_{U_3}^2(M_X)$, allowing heavy stop masses to cause fine
  tuning. In this case $m_{H_u}^2(M_{SUSY})$ is held fixed so as the
  cutoff increases the values of these softmasses at $M_X$ will be
  larger and there we be a bigger cancellation between them,
  increasing the sensitivity of $M_Z$ to both $m_{H_u}^2$ and the soft
  scalar masses for the stops.}.

@@
